Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-11-15 hazelnut-popl17.tex

\def\OPTIONConf{1}%
\newif\iftr\trtrue
\RequirePackage{etex}
\documentclass[9pt]{sigplanconf}
\toappear{}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage[lowtilde]{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 $^1$~Carnegie Mellon University (USA)
 \and
 $^2$~Oregon State University (USA)
 \and
 $^3$~University of Colorado Boulder (USA)\vspace{-10px}
}
{
%\email{comar@cs.cmu.edu}\\
%\texttt{http://www.cs.cmu.edu/\homedir comar/}
%\email{hiltonm@eecs.oregonstate.edu}\\
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

@@

Hazelnut is not intended as an end-user tool itself. Instead, it serves as a foundational account of typed structure
editing. To that end, we describe how
Hazelnut's rich metatheory, which we have mechanized using the Agda proof assistant, serves as a guide when we extend the calculus to include binary sum types. We also discuss various interpretations of holes, and in so doing
reveal connections with gradual typing and contextual modal type theory, the Curry-Howard interpretation of contextual modal logic. Finally, we
discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program. Our simple reference implementation is written in OCaml and compiled using \lstinline{js_of_ocaml}.

@@

\section{Related Work and Discussion}\label{sec:rw}
\todo{move into own file}
\subsection{Structure Editors}
Syntactic structure editors have a long history -- the Cornell Program
Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in
1981. Novice programmers have been a common target for structure
editors. For example, GNOME~\cite{garlan_gnome:_1984} was developed to
teach programming to undergraduates.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language
with an integrated structure editor for teaching novice CS undergraduate
students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure
editor targeted at children ages 8 to 16.
TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure
editor for programming on touch-based devices, and is used to teach high
school students. An implementation of Hazelnut might be useful in teaching
students about the typed lambda calculus, though that has not been our
explicit aim with this work.

@@

The examples given so far either define a language with a trivial static
semantics, or do not attempt to maintain well-typedness as an edit
invariant. This can pose problems, for reasons discussed in the
Introduction. One apparent exception is Unison~\cite{unison}, a structure
editor for a typed functional language similar to Haskell. Like Hazelnut,
it seems to define some notion of well-typedness for expressions with holes
(though there is no technical documentation on virtually any aspect of its
design.) Unlike Hazelnut, it does not have a notion analogous to Hazelnut's
notion of a non-empty hole. As such, programmers must construct programs in
a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another
system with the same problem is Polymorphic Blocks, a block-based user
interface where the structure of block connectors encodes a
type \cite{DBLP:conf/chi/LernerFG15}.

We fundamentally differ from these projects in our design philosophy: we
consider it essential to start by building type theoretic foundations,
which are independent of nearly all decisions about the user interface. In
contrast, these editors have developed innovative user interfaces (e.g. see
the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled
foundational calculus. In this respect, we follow the philosophical
approach taken by languages that are rooted in the type theoretic tradition
and have gone to great effort to develop a clear metatheory, like Standard
ML \cite{mthm97-for-dart,Harper00atype-theoretic}.  In the future, we hope
that these lines of research will merge to produce a human-usable typed
structure editor with sound formal foundations. Our contribution, then, is
in defining and analyzing the theoretical properties of a small
foundational calculus that could be extended to achieve this vision.

Some structure editor generators do operate on formal or semi-formal
definitions of an underlying language. For example, the Synthesizer
Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an
attribute grammar-based language implementation that then can be used to
generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005}
produces a language specific environment from a user defined formal
specification of a language. Barista is a programmatic toolkit for building
structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the
commercial JetBrains MPS framework for constructing structure
editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These
systems do not give a semantics to the edit states of the structure editor
itself, or maintain non-trivial edit invariants, as Hazelnut does.

@@

\subsection{Gradual Type Systems}
A significant contribution of this paper is the discovery of a clear
technical relationship between typed structure editing and gradual
typing. In particular, the machinery necessary to give a reasonable
semantics to type holes -- i.e. type consistency and type matching --
coincides with that developed in the study of gradual type systems for
functional languages. The pioneering work of Siek and Taha \cite{Siek06a}
introduced type consistency. Subsequent work developed the concept of type
matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In
retrospect, this relationship is perhaps unsurprising: gradual typing is,
notionally, also motivated by the idea of iterated development of a program
where every intermediate state is well-defined in some sense.

@@

An aspect of gradual typing that we did not touch on directly here is its
concern with assigning a dynamics to programs where type information is not
known, by inserting dynamic type casts. This would correspond to assigning
a dynamics to Hazelnut expressions with type holes such that a run-time
error occurs when a type hole is found to be unfillable through evaluation.

\subsection{Type Holes as Unification Variables}
Another possible interpretation of type holes is as explicit unification
variables. In other words, we might define the dynamics of Hazelnut such
that a program cannot be run if some automated type inference procedure
cannot statically fill in any type holes that arise. For a simple calculus,
e.g. the STLC upon which Hazelnut is based, type inference is known to be
decidable \cite{damas1982principal}. In more complex settings, e.g. in a
dependently typed language, a partial decision procedure may still be
useful in this regard, both at edit-time and (just prior to)
run-time. Indeed, text editor modes for proof assistants, e.g. for Agda,
attempt to do exactly this for indicated ``type holes'' (and do not always
succeed.)

\subsection{Exceptions}
Expression holes, too, could dynamically be interpreted in several
ways. One straightforward approach would be to dynamically raise an
exception if evaluation encounters an expression hole. Indeed,
placing \textt{raise Unimplemented} or similar in portions of an expression
that are under construction is a common practice across programming
languages.

\subsection{Type-Directed Program Synthesis}
Some text editor modes, e.g. those for proof assistants like
Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more
explicit hole-based programming model where indicated expression holes are
treated as sites where the system can be asked to automatically generate an
expression of an appropriate type. These systems are not statically
well-defined themselves (though see below.)

The Graphite system used an informal model of typed holes in Java to allow
library providers to associate interactive code generation interfaces with
types \cite{Omar:2012:ACC:2337223.2337324}.

@@

\subsection{Contextual Modal Type Theory}
Expression holes can also be understood by invoking the notion of
a \emph{metavariable} as found in contextual modal type theory
(CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}. In particular, expression
holes have types and are surrounded by contexts, just as metavariables in
CMTT are associated with types and contexts. This helps to clarify the
logical meaning of a typing derivation in Hazelnut -- it conveys
well-typedness relative to an (implicit) modal context that extracts each
expression hole's type and context. The modal context must be emptied --
i.e. the expression holes must be instantiated with expressions of the
proper type in the proper context -- before the expression can be
considered complete. This corresponds to the notion of modal necessity in
contextual modal logic.

Making the modal context explicit in our semantics is not technically
useful given our stated purpose -- interactive program editing is not
merely hole filling in Hazelnut (i.e. the cursor need not be on a
hole). Moreover, the hole's type and context become apparent as our action
semantics traverses the zipper structure on each step. Some interactive
proof assistants, however, support a tactic model based directly on hole
filling, so the connection to CMTT and similar systems is more useful. For
example, Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT
and aspects of Idris' editor support \cite{brady2013idris} are based on
McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have
explored a lambda calculus with contexts for a similar
purpose \cite{DBLP:conf/icfp/LeeF96}.

@@

\subsection{Evaluation Strategies: A High-Dimensional Space}
To summarize, we have discussed three different evaluation strategies in
the presence of type holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as unknown types, in the gradual sense.
\item ...as unification variables.
\end{enumerate}
In addition, we have discussed four different evaluation strategies in the
presence of expression holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as causing exceptions.
\item ...as sites for automatic program synthesis.
\item ...as the closures of CMTT.
\end{enumerate}

@@

\subsection{Future Work}
\subsubsection{Richer Languages}
Hazelnut is, obviously, a very limited language at its core. So the most
obvious avenue for future work is to increase the expressive power of this
language by extension. Our plan is to simultaneously maintain a
mechanization and implementation (following, for example, Standard ML) as
we proceed, ultimately producing the first large-scale, formally verified
bidirectionally typed structure editor.

@@

\subsubsection{Editor Services}
There are various aspects of the editor model that we have not yet
formalized. For example, our action model does not consider how actions are
actually entered using, for example, key combinations or chords. In
practice, we would want also to rank available actions in some reasonable
manner (perhaps based on usage data gathered from other users or code
repositories.) Designing a rigorous typed probability model over actions
and H-expressions is one avenue of research that we have started to
explore, with intriguing initial results.

\subsubsection{Programmable Actions}
Our language of actions is intentionally primitive. However, even now it
acts much like a simple imperative command language. This suggests future
expansion to, for example, a true tactic language. Alternatively, it may be
more useful to develop the notion of an \emph{action macro}, whereby
functional programs could themselves be lifted to the level of actions to
compute non-trivial compound actions. Such compound actions would give a
uniform description of transformations ranging from the simple -- like
``move the cursor to the next hole to the right'' -- to quite complex whole
program refactoring, while remaining subject to the core Hazelnut
metatheory.

@@

It should also be possible to develop a semantics of semantic comments,
i.e. comments that mention semantic structures. These would be subject to
the same operations, e.g. renaming, as other structures, helping to address
the problem of comments becoming inconsistent with code.

@@

\subsubsection{Empirical Evaluation}
Although we make few empirical claims in this paper, it is ultimately an
empirical question as to whether structure editors, and typed structure
editors, are practical. We hope to conduct user studies once a richer
semantics has been developed.

@@

\clearpage

\bibliographystyle{abbrvnat}
% The bibliography should be embedded for final submission.
\bibliography{bibliography}

\iftr
\clearpage
%\onecolumn
\appendix
\input{appendix}
\else
% No Appendix Here
\fi

>>>>

\def\OPTIONConf{1}%
\newif\iftr\trtrue
\RequirePackage{etex}
\documentclass[9pt]{sigplanconf}
\toappear{}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{balance}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage[lowtilde]{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 \begin{tabular}{ccc}
 	\begin{tabular}{c}
 $^1$Carnegie Mellon University (USA)\\
 \textsf{\hphantom{$^1$}\{comar, iev, aldrich\}@cs.cmu.edu}
 \end{tabular} & 
 \begin{tabular}{c}
 $^2$Oregon State University (USA)\\
 \textsf{\hphantom{$^1$}hiltonm@eecs.oregonstate.edu}
 \end{tabular} & 
  \begin{tabular}{c}
 $^3$University of Colorado Boulder (USA)\\
 \textsf{\hphantom{$^1$}matthew.hammer@colorado.edu}
 \end{tabular}
 \end{tabular}
}
{
 % \vspace{-12px}
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

@@

Hazelnut is not intended as an end-user tool itself. Instead, it serves as a foundational account of typed structure
editing. To that end, we describe how
Hazelnut's rich metatheory, which we have mechanized using the Agda proof assistant, serves as a guide when we extend the calculus to include binary sum types. We also discuss various interpretations of holes, and in so doing
reveal connections with gradual typing and contextual modal type theory, the Curry-Howard interpretation of contextual modal logic. Finally, we
discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program. Our simple reference implementation is written using \lstinline{js_of_ocaml}.

@@

\section{Related Work and Discussion}\label{sec:rw}
\subsection{Structure Editors}
Syntactic structure editors have a long history -- the Cornell Program
Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in
1981. Novice programmers have been a common target for structure
editors. For example, GNOME~\cite{garlan_gnome:_1984} was developed to
teach programming to undergraduates.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language
with an integrated structure editor for teaching novice CS undergraduate
students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure
editor targeted at children ages 8 to 16.
TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure
editor for programming on touch-based devices, and is used to teach high
school students. An implementation of Hazelnut might be useful in teaching
students about the typed lambda calculus, though that has not been our
explicit aim with this work.

@@

The examples given so far either do not attempt to reason statically about
types and binding, or do not attempt to maintain well-typedness as an edit
invariant. This can pose problems, for reasons discussed in Sec. \ref{sec:introduction}. One apparent exception is Unison~\cite{unison}, a structure
editor for a typed functional language similar to Haskell. Like Hazelnut,
it seems to define some notion of well-typedness for expressions with holes
(though there is no technical documentation on virtually any aspect of its
design.) Unlike Hazelnut, it does not have a notion analogous to Hazelnut's
notion of a non-empty hole. As such, programmers must construct programs in
a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another
system with the same problem is Polymorphic Blocks, a block-based user
interface where the structure of block connectors encodes a
type \cite{DBLP:conf/chi/LernerFG15}.

We fundamentally differ from these projects in our design philosophy: we
consider it essential to start by building type theoretic foundations,
which are independent of nearly all decisions about the user interface (other than our choice to use an explicit cursor.) In
contrast, these editors have developed innovative user interfaces (e.g. see
the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled
foundational calculus. In this respect, we follow the philosophical
approach taken by languages that are rooted in the type theoretic tradition
and have gone to great effort to develop a clear metatheory, like Standard
ML \cite{mthm97-for-dart,Harper00atype-theoretic,Lee:2007:TMM:1190216.1190245}.  In the future, we hope
that these lines of research will merge to produce a human-usable typed
structure editor with sound formal foundations. Our contribution, then, is
in defining and analyzing the theoretical properties of a small
foundational calculus that could be extended to achieve this vision. 
Our implementation resembles the minimal structure editor defined in 
Haskell by Sufrin and De Moor \cite{sufrin1999modeless}.

Some structure editor \emph{generators} do operate on formal or semi-formal
definitions of an underlying language. For example, the Synthesizer
Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an
attribute grammar-based language implementation that then can be used to
generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005}
produces a language specific environment from a user defined formal
specification of a language. Barista is a programmatic toolkit for building
structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the
commercial JetBrains MPS framework for constructing structure
editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These
systems do not give a semantics to the edit states of the structure editor
itself, or maintain non-trivial edit invariants, as Hazelnut does.

@@

Other work has attempted to integrate structure editing features into 
text editors. For example, recent work has used syntactic placeholders
reminiscent of our expression holes to decrease the percentage of 
edit states that are malformed \cite{Amorim:2016:PSC:2997364.2997374}. This work does not consider the semantics of placeholders.

Prior work has also explored 
formal definitions of text editor commands, e.g. using functional 
combinators \cite{DBLP:journals/scp/Sufrin82}.

\subsection{Gradual Type Systems}
A significant contribution of this paper is the discovery of a clear
technical relationship between typed structure editing and gradual
typing. In particular, the machinery necessary to give a reasonable
semantics to type holes -- i.e. type consistency and type matching --
coincides with that developed in the study of gradual type systems for
functional languages. The pioneering work of Siek and Taha \cite{Siek06a}
introduced type consistency. Subsequent work developed the concept of type
matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15} and has further 
studied the notion of type consistency \cite{Garcia:2016:AGT:2837614.2837670}. In
retrospect, this relationship is perhaps unsurprising: gradual typing is,
notionally, also motivated by the idea of iterated development of a program
where every intermediate state is well-defined in some sense, albeit at 
different granularity.

@@

An aspect of gradual typing that we did not touch on directly here is its
concern with assigning a dynamics to programs where type information is not
known, by inserting dynamic type casts \cite{Siek06a} or deducing evidence for consistency during evaluation \cite{Garcia:2016:AGT:2837614.2837670}. This would correspond to assigning
a dynamics to Hazelnut expressions with type holes such that a run-time
error occurs when a type hole is found to be unfillable through evaluation. This 
may be useful as an exploratory programming tool.

\subsection{Bidirectional Type Systems}
Hazelnut is bidirectionally typed \cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial,odersky2001colored}. Bidirectional type systems are notable in 
that they are easy to define, easy to implement, produce simple error messages and support advanced language features \cite{dunfield2013complete}. For example, Scala \cite{odersky2001colored} and Agda \cite{norell:thesis} are both fundamentally bidirectionally typed languages.

\subsection{Type Reconstruction}
An alternative approach to type inference is to use a unification-based type reconstruction system, as in functional languages like ML and Haskell \cite{damas1982principal}. This is difficult to reconcile with the approach presented in this paper, because edit actions could introduce new  unification constraints that would require placing non-empty holes around terms far from the cursor.  A whole-program hole insertion pass after each edit action could perhaps be used to recover invariants similar to those presented here, but we leave the details as future work. Our contention is that a bidirectional approach is a sweet spot in the design space of interactive systems like Hazelnut because it precludes
``spooky errors at a distance''. Instead, the interaction is a sort of local
dialog between the programmer and the system involving simple, familiar
concepts -- types with holes -- rather than sets of constraints.

An intermediate approach would be to 
layer unification-based type generation features atop the bidirectional system. This would amount to  interpretting type holes as unification
variables. For a simple calculus,
e.g. the STLC upon which Hazelnut is based, type inference for complete expressions is known to be
decidable, so type holes could be instantiated automatically once the expression that they appear within has been constructed. It would also be possible to flag expressions for which there does not exist any way to fill the type holes. In more complex settings, e.g. in a 
dependently typed language, a partial decision procedure may still be
useful in this regard, both at edit-time and (just prior to)
run-time. Indeed, text editor modes for dependently typed proof assistants, e.g. for Agda,
attempt to do exactly this for indicated ``type holes'' (and do not always
succeed.)

\subsection{Exceptions}
Expression holes can be interpreted in several
ways. One straightforward interpretation is to treat them 
like expressions that raise exceptions. Indeed,
placing \textt{raise Unimplemented} or similar in portions of an expression
that are under construction is a common practice across programming
languages today. The GHC dialect of Haskell recently introduced an explicit notion of a 
typed hole that behaves similarly \cite{GHC/holes}.

\subsection{Type-Directed Program Synthesis}
Some text editor modes, e.g. those for proof assistants like
Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more
explicit hole-based programming model where indicated expression holes are
treated as sites where the system can be asked to automatically generate an
expression of an appropriate type. %These systems are not statically
% well-defined themselves (though see below.)

The Graphite system borrowed Eclipse's heuristic model of typed holes for Java to allow
library providers to associate interactive code generation interfaces with
types \cite{Omar:2012:ACC:2337223.2337324}.

@@

\subsection{Tactics}

Interactive proof refinement systems, e.g. those in LCF \cite{Gordon:1978:MIP:512760.512773}, and more recent typed tactic systems, e.g. Mtac for Coq \cite{Ziliani:2013:MMT:2544174.2500579}, 
support an explicit model of a ``current'' typed hole that serves as the 
target of program synthesis. Hazelnut differs in that edits can occur anywhere within a term.

A related approach is to interpret expression holes as the \emph{metavariables} of contextual modal type theory
(CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}. In particular, expression
holes have types and are surrounded by contexts, just as metavariables in
CMTT are associated with types and contexts. This begins to clarify the
logical meaning of a typing derivation in Hazelnut -- it conveys
well-typedness relative to an (implicit) modal context that extracts each
expression hole's type and context. The modal context must be emptied --
i.e. the expression holes must be instantiated with expressions of the
proper type in the proper context -- before the expression can be
considered complete. This corresponds to the notion of modal necessity in
contextual modal logic.

We did not make the modal context explicit in our semantics because interactive program editing is not
merely hole filling in Hazelnut (i.e. the cursor need not be on a
hole.) Moreover, the hole's type and context become apparent as our action
semantics traverses the zipper structure on each step. For interactive
proof assistants that support a tactic model based directly on hole
filling, as just discussed, the connection to CMTT and similar systems is more useful. For
example, Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT
and aspects of Idris' editor support \cite{brady2013idris} are based on
McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have
explored a lambda calculus with contexts for a similar
purpose \cite{DBLP:conf/icfp/LeeF96}.

@@

@@

\subsection{Future Work}
\subsubsection{Richer Languages}
Hazelnut is, obviously, a very limited language at its core. So the most
obvious avenue for future work is to increase the expressive power of this
language by extension. Our plan is to simultaneously maintain a
mechanization and implementation (following, for example, Standard ML \cite{Lee:2007:TMM:1190216.1190245}) as
we proceed, ultimately producing the first large-scale, formally verified
bidirectionally typed structure editor.

@@

\subsubsection{Evaluation Strategies: A High-Dimensional Space}
The related work brought up in the previous section suggests three different evaluation strategies in
the presence of type holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as unknown types, in the gradual sense.
\item ...as unification variables.
\end{enumerate}
In addition, we have discussed four different evaluation strategies in the
presence of expression holes:
\begin{enumerate}[noitemsep]
\item ...as preventing evaluation (the standard approach.)
\item ...as causing exceptions.
\item ...as sites for automatic program synthesis.
\item ...as the closures of CMTT.
\end{enumerate}

\subsubsection{Editor Services}
There are various aspects of the editor model that we have not yet
formalized. For example, our action model does not consider how actions are
actually entered using, for example, key combinations or chords. In
practice, we would want also to suggest sensible compound actions, and 
to rank these suggested actions in some reasonable
manner (perhaps based on usage data gathered from other users or code
repositories.) Designing a action suggestion semantics and a rigorous typed probability model over actions is one avenue of research that we have started to
explore, with intriguing initial results.

\subsubsection{Programmable Actions}
Our language of actions is intentionally primitive. However, even now it
acts much like a simple imperative command language. This suggests future
expansion to, for example, a true \emph{action macro} language, whereby
functional programs could themselves be lifted to the level of actions to
compute non-trivial compound actions. Such compound actions would give a
uniform description of transformations ranging from the simple -- like
``move the cursor to the next hole to the right'' -- to quite complex whole
program refactoring, while remaining subject to the core Hazelnut
metatheory. Techniques like those in advanced tactic systems, e.g. Mtac, might be useful 
in proving these action macros correct \cite{Ziliani:2013:MMT:2544174.2500579}.

@@

It should also be possible to develop the concept of semantic comments,
i.e. comments that mention semantic structures or even contain values. These would be subject to
the same operations, e.g. renaming, as other structures, helping to address
the problem of comments becoming inconsistent with code. This system, generalized sufficiently,
could one day help unify document editing with program editing.

@@

\subsubsection{Empirical Evaluation}
Although we make few empirical claims in this paper, it is ultimately an
empirical question as to whether structure editors, and typed structure
editors, are practical. We hope to conduct user studies once a richer
semantics and a practical implementation thereof has been developed.

@@

\clearpage
\balance
\bibliographystyle{abbrvnat}
% The bibliography should be embedded for final submission.
\bibliography{bibliography}

% \iftr
% \clearpage
%\onecolumn
% \appendix
% \input{appendix}
% \else
% No Appendix Here
% \fi

### 2016-11-15 implementation.tex

% !TEX root = hazelnut-popl17.tex
\subsection{Implementation Concepts}
Central to any implementation of Hazelnut is a stream of edit states whose
cursor erasures synthesize types under an empty context according to the
synthetic action judgement,
$\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$. The middle
row of Figure \ref{fig:impl-overview} diagrams this stream of edit
states. By Theorem \ref{thrm:actsafe}, the editor does not need to
typecheck each edit state (though in some of the rules, e.g. Rule
(\ref{rule:zipper-asc}) which handles the situation where the type in a
type ascription changes, portions of the program do need to be
typechecked.) For example, the reader is encouraged to re-examine the
examples in Figure \ref{fig:first-example} and \ref{fig:second-example} --
the cursor erasure of each edit state synthesizes a type.

The programmer examines a view generated from each edit state and produces
actions in some implementation-defined manner (e.g. using a keyboard,
mouse, touchscreen, voice interface, or neural implant.) Each new action
causes a new abstract edit state to arise according to an implementation of
the action semantics. This then causes a new view to arise. This is a
simple event-based functional reactive programming model
\cite{Wan:2000:FRP:349299.349331}.

@@

The view computation renders the model (i.e. a Z-expression paired with a type) as nested HTML \texttt{div} elements matching the tree structure of the corresponding Z-expression. This tree is stylized separately using CSS. The action palette is a collection of buttons and text boxes which
are disabled when the corresponding action cannot be performed. We determine this by simply attempting to perform the
action internally and handling the exception that is raised when an action
is undefined. Each action form also has a corresponding keyboard shortcut. This design is not, of course, meant to score marks for
usability or performance. Instead, as a simple reference implementation, it allows the reader to interact with the system presented in this paper, to aid in their understanding.  Additionally, we expect its source code to be of use to others who are interested in
layering a more fluid user interface atop the core semantics, or extensions thereof.

>>>>

% !TEX root = hazelnut-popl17.tex
\subsection{Implementation Concepts}
Central to any implementation of Hazelnut is a stream of edit states whose
cursor erasures synthesize types under an empty context according to the
synthetic action judgement,
$\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$. The middle
row of Figure \ref{fig:impl-overview} diagrams this stream of edit
states. For example, the reader is encouraged to re-examine the
examples in Figure \ref{fig:first-example} and \ref{fig:second-example} --
the cursor erasure of each edit state synthesizes a type.

Because Theorem \ref{thrm:actsafe} expresses an invariant, the editor does not need to
typecheck the edit state anew on each action (though in some of the rules, e.g. Rule
(\ref{rule:zipper-asc}) which handles the situation where the type in a
type ascription changes, portions of the program would need to be
typechecked again.) In other words, many of the problems related to incrementality 
are simply not relevant. Storing the types of subtrees would allow for further
optimizations (e.g. of the zipper rules.)

The programmer examines a view generated from each edit state and produces
actions in some implementation-defined manner (e.g. using a keyboard,
mouse, touchscreen, voice interface, or neural interface), as diagrammed in Figure \ref{fig:impl-overview}. Each new action
causes a new abstract edit state to arise according to an implementation of
the action semantics. This then causes a new view to arise. This is a
simple event-based functional reactive programming model
\cite{Wan:2000:FRP:349299.349331}.

@@

The view computation renders the model (i.e. a Z-expression paired with a type) as nested HTML \texttt{div} elements matching the tree structure of the corresponding Z-expression. This tree is stylized separately using CSS. The action palette is a collection of buttons and text boxes which
are disabled when the corresponding action cannot be performed. We determine this by simply attempting to perform the
action internally and handling the exception that is raised when an action
is undefined. (An optimization that is not in HZ would be to implement a version of the action semantics that simply computes a boolean, rather than the resulting edit state, for the purposes of action validation.) Each action form also has a corresponding keyboard shortcut. For actions that take arguments, the keyboard shortcut moves the cursor into a text box. The action validation occurs on every change to the text box.

This implementation is not, of course, meant to score marks for
usability or performance (though both are surprisingly good for such a simple system.) Instead, as a simple reference implementation, it allows the reader to interact with the system presented in this paper, to aid in their understanding.  Additionally, we expect its source code to be of use to others who are interested in
layering a more fluid user interface atop the core semantics, or extensions thereof.
